\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2017-10-11 paper/casebase_jss.Rmd
condense case study 1

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
library(tidyverse)
library(magrittr)
```

@@

  - Motivation
    + Flexible 
    + Flexible
    + Flexible

@@

@@

# Case study 1: Veteran data (or ERSPC if we can)

  - First example
  - Show how we can test for non-proportional hazard?

# Case study 2: Bone-marrow transplant

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
library(splines)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)
```

@@

The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient. The following points should be noted:

1. Time matching/risk set sampling (including Cox partial likelihood) eliminates the baseline hazard from the likelihood expression for the
hazard ratios
2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]

[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression. From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.

@@

The `casebase` package fits the family of hazard functions of the form

$$ h(x,t) = exp[g(x,t)] $$
where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any). Different functions of $t$ lead to different parametric hazard models.

The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.

$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$

The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).

The Gompertz hazard model is given by including a linear term for time:

$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$

Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$

@@

# Case study 1--European Randomized Study of Prostate Cancer Screening

To introduce the different features available, we make use of the European Randomized Study of Prostate Cancer Screening data; this dataset is available through the \pkg{casebase} package:

```{r}
data(ERSPC)
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
```

The results of this study were published by [schroder2009screening]. This data was obtained using the approach described in [liu2014recovering].

Population time plots can be extremely informative graphical displays of survival data. They should be the first step in an exploratory data analysis. We facilitate this task in the \pkg{casebase} package using the \code{popTime} function. We first create the necessary dataset for producing the population time plots, and we can produce the plot by using the corresponding \code{plot} method:

```{r}
pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
plot(pt_object)
```

We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:

```{r}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
plot(pt_object_strat)
```

We can also plot them side-by-side using the \code{ncol} argument:

```{r}
plot(pt_object_strat, ncol = 2)
```

First, we fit a Cox model to the data, examine the hazard ratio for the screening group (relative to the control group), and plot the cumulative incidence function (CIF).

```{r}
cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                             data = ERSPC)
summary(cox_model)
```

We can plot the CIF for each group:

```{r}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
                       ignore = 99)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]))
legend("topleft", 
       legend = c("Control group", "Screening group"), 
       col = c("red","blue"),
       lty = c(1, 1), 
       bg = "gray90")
```

Next we fit several models using case-base sampling. The models we fit differ in how we choose to model time.

The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).

```{r}
casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
                                                  data = ERSPC, 
                                                  ratio = 100)

summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2,])
```

The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation:

$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$

In the plot below, we overlay the estimated CIF from the casebase exponential model on the Cox model CIF:

```{r}
smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]))
lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

As we can see, the exponential model is not a good fit. Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit is expected. A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot. This example demonstrates the benefits of population time plots as an exploratory analysis tool.

Next we enter time linearly into the model:

```{r}
casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
                                 data = ERSPC, 
                                 ratio = 100)

summary(casebase_time)
exp(coef(casebase_time))
exp(confint(casebase_time))
```

```{r}
smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
                                          time = seq(0,15,0.1), 
                                          newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]))
lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

We see that the Weibull model leads to a better fit.

Next we try to enter a smooth function of time into the model using the \code{splines} package:

```{r}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
                                    data = ERSPC, 
                                    ratio = 100)

summary(casebase_splines)
exp(coef(casebase_splines))
exp(confint(casebase_splines))
```

```{r}
smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
                                    time = seq(0,15,0.1), 
                                    newdata = new_data)

plot(survfit(cox_model, newdata = new_data),
     xlab = "Years since Randomization", 
     ylab = "Cumulative Incidence", 
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
                    exp(coef(cox_model)), 
                    exp(confint(cox_model))[1], 
                    exp(confint(cox_model))[2]))
lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)

legend("topleft", 
       legend = c("Control group (Cox)","Control group (Casebase)",
                  "Screening group (Cox)", "Screening group (Casebase)"), 
       col = c("red","red", "blue","blue"),
       lty = c(1, 2, 1, 2), 
       bg = "gray90")
```

It looks like the best fit.

Since we are within the GLM framework, we can easily test for which model better fits the data using a Likelihood Ratio Test (LRT). The null hypothesis here is that the linear model is just as good as the larger (in terms of number of parameters) splines model.

```{r}
anova(casebase_time, casebase_splines, test = "LRT")
```

As espected, we see that splines model provides a better fit.

# Case study 2--Bone-marrow transplant

@@

In the following table we provide a comparison between the Cox model and case-base sampling:

```{r, echo=FALSE, eval = FALSE, results='asis'}
knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
```

### 2017-10-11 paper/casebase_jss.tex
condense case study 1

\begin{itemize}
\tightlist
\item
  Motivation

  \begin{itemize}
  \tightlist
  \item
    Flexible
  \item
    Flexible
  \item
    Flexible
  \end{itemize}
\end{itemize}

@@

@@

\section{Case study 1: Veteran data (or ERSPC if we
can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}

\begin{itemize}
\tightlist
\item
  First example
\item
  Show how we can test for non-proportional hazard?
\end{itemize}

\section{Case study 2: Bone-marrow
transplant}\label{case-study-2-bone-marrow-transplant}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.82 & (0.46, 1.48)\\
\hline
Disease & 0.48 & (0.25, 0.92)\\
\hline
Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
\hline
Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
\hline
Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
\hline
Source & 1.51 & (0.49, 4.69)\\
\hline
Age & 0.99 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}

@@

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
\end{figure}
\end{CodeChunk}

\section{Case study 3: Vaccination study (recurrent
events)}\label{case-study-3-vaccination-study-recurrent-events}

\begin{itemize}
\tightlist
\item
  Give a more complex example of sampling; time-dependent exposure

  \begin{itemize}
  \tightlist
  \item
    Sampling needs to be done manually, but fitting function can still
    be used
  \end{itemize}
\end{itemize}

\renewcommand\refname{Discussion}
\bibliography{casebase_references.bib}

>>>>

The purpose of the \texttt{casebase} package is to provide practitioners
with an easy-to-use software tool to predict the risk (or cumulative
incidence (CI)) of an event, for a particular patient. The following
points should be noted:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Time matching/risk set sampling (including Cox partial likelihood)
  eliminates the baseline hazard from the likelihood expression for the
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \citep{hanley2009fitting}
\end{enumerate}

\citep{hanley2009fitting} propose a fully parametric hazard model that
can be fit via logistic regression. From the fitted hazard function,
cumulative incidence and, thus, risk functions of time, treatment and
profile can be easily derived.

@@

The \texttt{casebase} package fits the family of hazard functions of the
form

\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
(number of units) of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any). Different functions of \(t\) lead to
different parametric hazard models.

The simplest of these models is the one-parameter exponential
distribution which is obtained by taking the hazard function to be
constant over the range of \(t\).

\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been on study a.k.a
the memoryless property (Kalbfleisch and Prentice, 2002).

The Gompertz hazard model is given by including a linear term for time:

\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]

Use of \(log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]

@@

\section{Case study 1--European Randomized Study of Prostate Cancer
Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}

To introduce the different features available, we make use of the
European Randomized Study of Prostate Cancer Screening data; this
dataset is available through the \pkg{casebase} package:

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0,1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by {[}schroder2009screening{]}.
This data was obtained using the approach described in
{[}liu2014recovering{]}.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in an exploratory data
analysis. We facilitate this task in the \pkg{casebase} package using
the \code{popTime} function. We first create the necessary dataset for
producing the population time plots, and we can produce the plot by
using the corresponding \code{plot} method:

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> plot(pt_object)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-2-1} \end{center}

We can also create exposure stratified plots by specifying the
\code{exposure} argument in the \code{popTime} function:

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R+                                      event = "DeadOfPrCa", 
R+                                      exposure = "ScrArm")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> plot(pt_object_strat)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-3-1} \end{center}

We can also plot them side-by-side using the \code{ncol} argument:

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-4-1} \end{center}

First, we fit a Cox model to the data, examine the hazard ratio for the
screening group (relative to the control group), and plot the cumulative
incidence function (CIF).

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
Call:
survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
    ScrArm, data = ERSPC)

  n= 159893, number of events= 540

                        coef exp(coef) se(coef)     z Pr(>|z|)  
ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

                      exp(coef) exp(-coef) lower .95 upper .95
ScrArmScreening group     0.801       1.25     0.674     0.952

Concordance= 0.519  (se = 0.011 )
Rsquare= 0   (max possible= 0.075 )
Likelihood ratio test= 6.45  on 1 df,   p=0.0111
Wald test            = 6.37  on 1 df,   p=0.0116
Score (logrank) test = 6.39  on 1 df,   p=0.0115
\end{CodeOutput}
\end{CodeChunk}

We can plot the CIF for each group:

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1} \end{center}

Next we fit several models using case-base sampling. The models we fit
differ in how we choose to model time.

The \code{fitSmoothHazard} function provides an estimate of the hazard
function \(h(x, t)\) is the hazard function, where \(t\) denotes the
numerical value of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any).

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R+                                                   data = ERSPC, 
R+                                                   ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-0.458  -0.458  -0.409  -0.409   2.247

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -7.7546     0.0584  -132.8 <0.0000000000000002 ***
ScrArmScreening group  -0.2397     0.0921    -2.6              0.0093 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3612.2  on 5938  degrees of freedom
AIC: 3616

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
ScrArmScreening group 
                 0.79 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
 2.5 % 97.5 % 
  0.66   0.94 
\end{CodeOutput}
\end{CodeChunk}

The \code{absoluteRisk} function provides an estimate of the cumulative
incidence curves for a specific risk profile using the following
equation:

\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]

In the plot below, we overlay the estimated CIF from the casebase
exponential model on the Cox model CIF:

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-8-1} \end{center}

As we can see, the exponential model is not a good fit. Based on what we
observed in the population time plot, where more events are observed
later on in time, this poor fit is expected. A constant hazard model
would overestimate the cumulative incidence earlier on in time, and
underestimate it later on; this is what we see on the cumulative
incidence plot. This example demonstrates the benefits of population
time plots as an exploratory analysis tool.

Next we enter time linearly into the model:

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R+                                  data = ERSPC, 
R+                                  ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.030  -0.489  -0.367  -0.278   2.740

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -9.0228     0.1186  -76.07 <0.0000000000000002 ***
Follow.Up.Time          0.2217     0.0159   13.97 <0.0000000000000002 ***
ScrArmScreening group  -0.2680     0.0939   -2.85              0.0043 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3402.7  on 5937  degrees of freedom
AIC: 3409

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
          (Intercept)        Follow.Up.Time ScrArmScreening group 
              0.00012               1.24824               0.76490 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                         2.5 %  97.5 %
(Intercept)           0.000095 0.00015
Follow.Up.Time        1.210257 1.28799
ScrArmScreening group 0.635606 0.91877
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-10-1} \end{center}

We see that the Weibull model leads to a better fit.

Next we try to enter a smooth function of time into the model using the
\code{splines} package:

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
R+                                     data = ERSPC, 
R+                                     ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.245  -0.521  -0.397  -0.211   3.208

Coefficients:
                             Estimate Std. Error z value
(Intercept)                  -10.5377     0.3384  -31.14
splines::bs(Follow.Up.Time)1   4.9852     0.8551    5.83
splines::bs(Follow.Up.Time)2   1.7658     0.4963    3.56
splines::bs(Follow.Up.Time)3   5.1969     0.7203    7.22
ScrArmScreening group         -0.1850     0.0941   -1.97
                                         Pr(>|z|)    
(Intercept)                  < 0.0000000000000002 ***
splines::bs(Follow.Up.Time)1     0.00000000553602 ***
splines::bs(Follow.Up.Time)2              0.00037 ***
splines::bs(Follow.Up.Time)3     0.00000000000054 ***
ScrArmScreening group                     0.04937 *  
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3352.4  on 5935  degrees of freedom
AIC: 3362

Number of Fisher Scoring iterations: 6
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
                 (Intercept) splines::bs(Follow.Up.Time)1 
                    0.000027                   146.229029 
splines::bs(Follow.Up.Time)2 splines::bs(Follow.Up.Time)3 
                    5.846329                   180.718284 
       ScrArmScreening group 
                    0.831083 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                                 2.5 %    97.5 %
(Intercept)                   0.000013   0.00005
splines::bs(Follow.Up.Time)1 28.719353 825.32637
splines::bs(Follow.Up.Time)2  2.210676  15.57574
splines::bs(Follow.Up.Time)3 44.370102 759.24168
ScrArmScreening group         0.690354   0.99867
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-12-1} \end{center}

It looks like the best fit.

Since we are within the GLM framework, we can easily test for which
model better fits the data using a Likelihood Ratio Test (LRT). The null
hypothesis here is that the linear model is just as good as the larger
(in terms of number of parameters) splines model.

\begin{CodeInput}
R> anova(casebase_time, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
Analysis of Deviance Table

Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
Model 2: DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm + offset(offset)
  Resid. Df Resid. Dev Df Deviance       Pr(>Chi)    
1      5937       3403                               
2      5935       3352  2     50.3 0.000000000012 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

As espected, we see that splines model provides a better fit.

\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.72 & (0.4, 1.29)\\
\hline
Disease & 0.61 & (0.33, 1.15)\\
\hline
Phase (CR2 vs. CR1) & 1.31 & (0.52, 3.31)\\
\hline
Phase (CR3 vs. CR1) & 1.58 & (0.39, 6.32)\\
\hline
Phase (Relapse vs. CR1) & 4.50 & (2.07, 9.77)\\
\hline
Source & 1.42 & (0.46, 4.41)\\
\hline
Age & 1.00 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-19-1}

@@

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-19}
\end{figure}
\end{CodeChunk}

\section{Case study 3}\label{case-study-3}

\begin{itemize}
\tightlist
\item
  Glmnet, gam, and gbm
\end{itemize}

\section{Discussion}\label{discussion}

In the following table we provide a comparison between the Cox model and
case-base sampling:

\bibliography{casebase_references.bib}
